Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-07-28 paper-amia-summit-2018/manuscript.tex
update methods section

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/segment-example.png}
    \caption{\textbf{Example of e-Coaching emails segmented into fragments that correspond to MI behaviors of an e-Coach and a patient}.}
    \label{fig:text-segment}
\end{figure}

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three neural networks models. Since neural networks provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.\\

@@

\textbf{Conditional Random Field}: state-of-the-art supervised classification method proven to be effective for text categorization\cite{joachims1998text} for its ability to cope with high dimensional input feature space. SVM finds the hyperplane that maximizes the separation between the closest ``new segment'' and ``same segment'' training examples. We scaled our dataset to utilize linear SVM with optimal cost 512 in experiments.

\textbf{Multilayer Perceptrons}: this method considers each training sample as a point in the input feature space. For a new test sample, Euclidean distance is calculated to find the k-nearest classified neighbors from the training set. Finally, the test sample is classified into a majority class of its k-nearest neighbors. We experimentally determined that best performance was achieved when the number of neighbors is 3.

\textbf{Bidirectional Recurrent Neural Networks}: RNN is a neural network architecture designed to capture sequential patterns. When we predict the new segment point using RNNs, particular combinations of words in a sequence will indicate the change of MI behavior. Long Short Term Memory (LSTM) networks\cite{hochreiter1997long} are a special type of RNN that are capable of handling variable size input sequence and have an internal memory that can be reset. GRU\cite{chung2014empirical} is a variant of LSTM mathematically represented by the following formula:

@@

\textbf{Convolutional Recurrent Neural Networks}: RNN is a neural network architecture designed to capture sequential patterns. When we predict the new segment point using RNNs, particular combinations of words in a sequence will indicate the change of MI behavior. Long Short Term Memory (LSTM) networks\cite{hochreiter1997long} are a special type of RNN that are capable of handling variable size input sequence and have an internal memory that can be reset. GRU\cite{chung2014empirical} is a variant of LSTM mathematically represented by the following formula:

Our GRU model includes one hidden layer, output layer and input layer. We reset our model state after feeding each input sequence where input was given as one-hot encoding of word vector. Since LSTM and GRU use one-hot vector as input, results for these models are reported only when lexical and punctuation features are used. We experimentally determined that the best performance is achieved when the number of hidden units is 25, batch size is 1 and Adam\cite{kingma2014adam} is used for optimization. Source codes of this work will be available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication.

@@

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, SVM has the best performance across all topic model-based features in terms of all performance metrics, achieves 0.990 precision, recall and F1-measure when DL-LDA-based topic features are used. Similar to results in Tables~\ref{tab:result_base}, \ref{tab:result_boundary} and \ref{tab:result_not_boundary}, NB demonstrates the lowest performance among all methods while best performance achieves with LCA-based topic features compared to other model-based topic features. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.9\% and 0.2\%; recall increases by 0.9\% and 0.2\%; and F1-measure increases by 0.9\% and 0.3\% for SVM and KNN methods, respectively, when lexical feature is used in combination with punctuation and best model-based topic features.\\

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall and F1-measure were used to eliminate doubt about the model performance since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. Although NB provides lowest classification results, it shows moderate AUC values, achieving 0.978 AUC when only lexical features are used and 0.977 AUC when a combination of lexical, punctuation and topic features are used. On the other hand, KNN and SVM exhibit 0.972 and 0.986 AUCs when only lexical features are used; and 0.966 and 0.980 AUCs when a combination of lexical, punctuation and topic model-based features are used. Finally, LSTM demonstrates the highest AUC value among all machine learning models, achieving 0.996 AUC, while GRU achieves only 0.2\% lower AUC than LSTM.

>>>>

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/segment-example.png}
    \caption{\textbf{Example of e-Coaching emails segmented into fragments that correspond to MI behaviors of an e-Coach and a patient}.}
    \label{fig:text-segment}
\end{figure}

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three neural networks models. Since neural networks provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Fields}: CRF have been widely used in various NLP tasks including part-of-speech tagging and segmentation tasks \cite{lafferty2001conditional, hirohata2008identifying}. Unlike a maximum entropy markov model which uses per-state exponential models for the conditional probabilities of next states given the current state, a CRF is a discriminative model of the conditional distribution of the entire sequence of labels given the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ for output and input sequences, $y$ and $x$, by the following equation:

\begin{equation}
p(y|x) = \frac{1}{Z_x}\exp{\left(\sum_{t=1}^{T}\sum_{k}^{}\lambda_k f_k(y_{t-1}, y_t, x, t)\right)} \\
\end{equation}

Here $Z_x$ is a normalization factor over all possible labelings of $x$, and $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently by the Viterbi algorithm. In our experiments, the following features were utilized with CRF models: i) current word/punctuation ii) next and previous 3 words/punctuations iii) whether the word/punctuation is special characters (';', '?', '.', ',', '!', ':', etc) or not iv) whether the word is title or not? (e.g. "The" is a title word but "the" is not) v) POS tags

\textbf{Multilayer Perceptrons}: MLP is a feed forward artificial neural network model which maps set of input onto a set of appropriate output. This study utilized MLP with a nonlinear activation function (relu) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we implement dropout on fully connected layers, which randomly hide some neurons during the training phase \cite{srivastava2014dropout}. Dropout is also applied in the fully connected layers in BRNN and CRNN models. For a given sample with $2l$ words/punctuations, the first layer input of MLP was created by the summation of first $l$ word vectors and last $l$ word vectors.

@@

\textbf{Bidirectional Recurrent Neural Networks}: BRNN is a neural network designed to capture sequential patterns by considering both past and future inputs as well as complex relationships between input feature and output labels. The output of BRNN layer is computed as the summation of the forward RNN output with backward RNN output. Gated Recurrent Units (GRU) \cite{chung2014empirical} are a special type of RNN that are capable of handling variable size input sequence and have an internal memory that can be reset. Figure~\ref{fig:crnn} will represent an architecture of BRNN if convolution layer is removed.

\textbf{Convolutional Recurrent Neural Networks}: CRNN is a deep neural network architecture consisting of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output.

@@

A sequence of $m$ words is fed into the input layer which provides a $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embedding, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = n_w + n_p$ of size 310. The 1D convolutional layer generates $m - h_c + 1$ new features with $h_c.n_e$ a filter of $h_c$ words and zero-padding on both sides of the text. A pooling operation was performed over time to find the most significant features in a region. After the convolutional layer, $m \times n_f$ are fed into bidirectional recurrent neural networks with 200 hidden units. Dropout is used to randomly drop 50\% of the neurons and passes them through a fully connected layer, where a sigmoid operation is calculated, giving us the probability of whether the word precedes a new segment or not. We experimentally determined that the best performance is achieved when filter length in convolution layer is 7, pool size is 3, batch size is 32, adam\cite{kingma2014adam} is used for optimization and the early stopping strategy is used with dropout to prevent over-fitting. Source codes of all experiments will be available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication.

@@

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, SVM has the best performance across all topic model-based features in terms of all performance metrics, achieves 0.990 precision, recall and F1-measure when DL-LDA-based topic features are used. Similar to results in Tables~\ref{tab:result_base}, \ref{tab:result_boundary} and \ref{tab:result_not_boundary}, NB demonstrates the lowest performance among all methods while best performance achieves with LCA-based topic features compared to other model-based topic features. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.9\% and 0.2\%; recall increases by 0.9\% and 0.2\%; and F1-measure increases by 0.9\% and 0.3\% for SVM and KNN methods, respectively, when lexical feature is used in combination with punctuation and best model-based topic features.

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall and F1-measure were used to eliminate doubt about the model performance since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. Although NB provides lowest classification results, it shows moderate AUC values, achieving 0.978 AUC when only lexical features are used and 0.977 AUC when a combination of lexical, punctuation and topic features are used. On the other hand, KNN and SVM exhibit 0.972 and 0.986 AUCs when only lexical features are used; and 0.966 and 0.980 AUCs when a combination of lexical, punctuation and topic model-based features are used. Finally, LSTM demonstrates the highest AUC value among all machine learning models, achieving 0.996 AUC, while GRU achieves only 0.2\% lower AUC than LSTM. \\
